The role of advection versus diffusion in intraventricular transport

### 2025-03-18 paper/main.tex
update peclet# values, caption fig1

We model the net forces induced by motile cilia as a steady
bidirectional traction acting on parts of the ventricular wall
(\Crefpanels{fig:fig1}{d}). Simulations show that this traction alone
leads to a partial compartmentalization of the ventricular system,
with large-scale vortex structures in the anterior, middle and
posterior ventricles (\Crefpanels{fig:fig1}{e}). The highest flow
speeds ($26.5$ \textmu m/s) occur in the vicinity of the dorsal cilia
in the middle ventricle. To model the cardiac pulsations, we impose a
pulsatile pressure difference between the anterior and posterior
boundaries (\Crefpanels{fig:fig1}{d, \fixme{g}}). The resulting flow
is pulsatile and directional (\Crefpanels{fig:fig1}{\fixme{f}}), and
in the absence of the cilia, the vortex structures vanish. The
velocity magnitude peaks at $4.8$ \textmu m/s. CSF initially flows
rostrocaudally before changing direction to caudorostral in the middle
of the cardiac cycle.

When combining the contributions, we recognize flow features
persisting throughout the cardiac cycle, now with vortex structures in
the anterior and middle ventricles and directional flow in the
posterior ventricle (\Crefpanels{fig:fig1}{h--j}, \fixme{Supplementary
  video 1}). These characteristic patterns suggest that the flow in
the anterior and middle ventricles is primarily driven by the motile
cilia, while the cardiac pressure pulsations dominate within the
posterior ventricle and in the ducts connecting the ventricular
compartments. The CSF flow speed peaks at the time of peak caudal
cardiac pulsatile flow, when both the tangential traction induced by
the cilia and the pulsatile pressure-driven flow are aligned with the
rostrocaudal axis, reaching up to 27.7 \textmu m/s dorsally in the
posterior part of the middle ventricle. The flow Reynolds number is
$\mathrm{Re} \approx 0.004$, justifying the assumption of Stokes
flow. In summary, the computational CSF flow model reproduces the flow
features observed in vivo~\cite{Olstad2019CiliaryDevelopment}.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{graphics/figure1.png}
    \caption{
    \textbf{a.} Schematic illustration of the two cerebrospinal fluid (CSF) flow components modeled,
    motile cilia and cardiac pulsations, and their contributions to the flow patterns.
    \textbf{b, c.} Particle tracking data visualizing CSF flow directionality (\textbf{b}) and
    velocity magnitudes (\textbf{c}) in the middle ventricle.
    \textbf{d.} Computational mesh with marked facets indicating regions of motile cilia (magenta) and
    anterior/posterior facets (blue).
    \textbf{e.} Streamlines of the (steady-state) CSF flow field simulated with the
    cilia-only/no-cardiac flow model.
    \textbf{f.} Streamlines of the CSF flow field simulated with the cardiac-only/no-cilia flow
    model at a time 25\% into the cardiac cycle. %t=0.112613 s, period=0.45045 s
    \textbf{g.} Illustration of the cardiac-related normal pressure forcing imposed on the
    anterior/posterior boundaries.
    \textbf{h--j.} Streamlines simulated with the baseline flow model at
    a time a time 25\% (\textbf{h}), 50\% (\textbf{i}) and 75\% (\textbf{j})
    into the cardiac cycle. Times indicated in panel \textbf{g}.
    %t=0.112613 s %t=0.225225 s %t=0.315315 s
    Scale bars 50 \textmu m.
    }
    \label{fig:fig1}
\end{figure}

%% Over one cardiac cycle, the CSF flow speed reaches up to
%% 27.7 \textmu m/s, dorsally in the posterior part of the middle
%% ventricle. This peak is attained at the time instant coinciding with
%% the time of peak caudal cardiac pulsatile flow,

@@

Interestingly, the distribution patterns differ between the
solutes. For the smallest diffusion coefficient, representing
extracellular vesicles, the advective transport by the CSF dominates diffusion: the
CSF flow structures carry over into the concentration field,
indicating that transport along the streamlines is more rapid than the
diffusion across the streamlines (\Crefpanels{fig:fig2}{b}). The
balance between diffusion and advection shifts with increasing
diffusion coefficient as the solute spreads more uniformly throughout
the ventricular geometry (\Crefpanels{fig:fig2}{b--d}). Notably, the
smaller solutes with higher diffusion coefficients spread more quickly
to distal regions. Computing Péclet numbers, we find that
$\mathrm{Pe}_1=519$, $\mathrm{Pe}_2=14.7$ and $\mathrm{Pe}_3=7.4$ for
$D_1$, $D_2$ and $D_3$, respectively. With all $\mathrm{Pe} > 1$, this
indicates that advection is the dominant transport mechanism on a
global scale.

@@

The CSF velocities are most strongly
affected when removing cilia in the dorsal diencephalic ventricle,
as the maximum velocity magnitude drops from 27.7 \textmu m/s to 8.0 \textmu m/s.
In vivo, these cilia generate the strongest flow at
this developmental stage of the fish~\cite{Olstad2019CiliaryDevelopment}.
In contrast, removing telencephalic and ventral diencephalic cilia has minimal
impact and results in maximum speeds of 27.7 \textmu m/s or 27.2 \textmu m/s, respectively.
Of the total tangential forces exerted by the cilia, ventral cilia in the middle ventricle
contribute with 34\%, while the dorsal cilia exert 52\% of the total tangential
force~(\Crefpanels{fig:fig5}{b}). The magnitude of the force contributions are thus
more similar than the difference in their impact on the maximum velocity magnitudes.

@@

Another study that used a force density to model ciliary motion is
Thouvenin~\emph{et al.}~\cite{Thouvenin2020OriginCanal}, where two-dimensional flow
in the central canal of zebrafish embryos was modeled using a cilia force density of
4000 N/$\mathrm{m^2}$. The resulting numerical velocity fields matched well with
in vivo measurements, where the maximum velocity magnitude was around 5 \textmu m/s
(recall that our model exhibits a maximum of 27.7 \textmu m/s).
The geometry they studied resembles a pipe, probably offering more flow resistance
than the open geometry considered in our work, which may be why they needed a larger
force density to accurately represent experimental results. In any case, it is
interesting to note that our and these two previous studies have estimated
cilia force densities that only differ by one order of magnitude.

@@

To evaluate the relative importance of advection and
diffusion as transport mechanisms, we consider the Péclet number
\begin{equation*}
    \mathrm{Pe} = \frac{\mathrm{diffusion \ timescale}}{\mathrm{advection \ timescale}}
    = \frac{t_d}{t_a} = \frac{L_p^2/D}{L_p/U_p} = \frac{L_p U_p}{D}.
\end{equation*}
Thus, if $\mathrm{Pe} > 1$, transport is dominated by advection,
and if $\mathrm{Pe} < 1$ transport is dominated by diffusion.
Here, $L_p$ and $U_p$ are characteristic length and velocity scales,
respectively, and $D$ is the solute diffusion coefficient.
In calculating a global Péclet number, we use $L_p=600$ \textmu m, the approximate length of the
ventricular geometry along the rostrocaudal ($x$-) axis.
We use a mean velocity $\overline{U}$ as the characteristic velocity $U_p$.
The mean velocity $\overline{U}$ is calculated by first averaging the velocity in space:
\begin{equation*}
    \overline{u}^2 = \frac{1}{|\Omega|}\int_{\Omega}\uu\cdot\uu\dx,
\end{equation*}
and then averaging $\overline{u}$ in time over one cardiac cycle to yield $\overline{U} = 2.4$ \textmu m/s.
To assess the assumption that inertial forces are negligible compared to
viscous forces, we also compute the Reynolds number
\begin{equation*}
    \mathrm{Re} = \frac{\mathrm{inertial \ forces}}{\mathrm{viscous \ forces}} = \frac{\rho U_r L_r}{\mu},
\end{equation*}
with $L_r = 110$ \textmu m (the height of the middle ventricle) and
$U_r = 27.7$ \textmu m/s (the maximum velocity magnitude)
as length and velocity scales characteristic of the flow.

>>>>

We model the net forces induced by motile cilia as a steady
bidirectional traction acting on parts of the ventricular wall
(\Crefpanels{fig:fig1}{d}). Simulations show that this traction alone
leads to a partial compartmentalization of the ventricular system,
with large-scale vortex structures in the anterior, middle and
posterior ventricles (\Crefpanels{fig:fig1}{e}). The highest flow
speeds ($\fixme{26.5}$ \textmu m/s) occur in the vicinity of the dorsal cilia
in the middle ventricle. To model the cardiac pulsations, we impose a
pulsatile pressure difference between the anterior and posterior
boundaries (\Crefpanels{fig:fig1}{d, \fixme{g}}). The resulting flow
is pulsatile and directional (\Crefpanels{fig:fig1}{\fixme{f}}), and
in the absence of the cilia, the vortex structures vanish. The
velocity magnitude peaks at $\fixme{4.8}$ \textmu m/s. CSF initially flows
rostrocaudally before changing direction to caudorostral in the middle
of the cardiac cycle.

When combining the contributions, we recognize flow features
persisting throughout the cardiac cycle, now with vortex structures in
the anterior and middle ventricles and directional flow in the
posterior ventricle (\Crefpanels{fig:fig1}{h--j}, \fixme{Supplementary
  video 1}). These characteristic patterns suggest that the flow in
the anterior and middle ventricles is primarily driven by the motile
cilia, while the cardiac pressure pulsations dominate within the
posterior ventricle and in the ducts connecting the ventricular
compartments. The CSF flow speed peaks at the time of peak caudal
cardiac pulsatile flow, when both the tangential traction induced by
the cilia and the pulsatile pressure-driven flow are aligned with the
rostrocaudal axis, reaching up to 28.0 \textmu m/s dorsally in the
posterior part of the middle ventricle. The flow Reynolds number is
$\mathrm{Re} \approx 0.004$, justifying the assumption of Stokes
flow. In summary, the computational CSF flow model reproduces the flow
features observed in vivo~\cite{Olstad2019CiliaryDevelopment}.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{graphics/figure1.png}
    \caption{
    \textbf{a.} Schematic illustration of the two cerebrospinal fluid (CSF) flow components modeled,
    motile cilia and cardiac pulsations, and their contributions to the flow patterns.
    \textbf{b, c.} Particle tracking data visualizing CSF flow directionality (\textbf{b}) and
    velocity magnitudes (\textbf{c}) in the middle ventricle.
    \textbf{d.} Computational mesh with marked facets indicating regions of motile cilia (magenta) and
    anterior/posterior facets (blue). Dorsal (D), ventral (V), anterior (A), posterior (P),
    left (L) and right (R) coordinates indicated.
    \textbf{e.} Streamlines of the (steady-state) CSF flow field simulated with the
    cilia-only/no-cardiac flow model.
    \textbf{f.} Streamlines of the CSF flow field simulated with the cardiac-only/no-cilia flow
    model at a time 25\% into the cardiac cycle. %t=0.112613 s, period=0.45045 s
    \textbf{g.} Illustration of the cardiac-related normal pressure forcing imposed on the
    anterior/posterior boundaries.
    \textbf{h--j.} Streamlines simulated with the baseline flow model at
    a time a time 25\% (\textbf{h}), 50\% (\textbf{i}) and 75\% (\textbf{j})
    into the cardiac cycle. Times indicated in panel \textbf{g}.
    %t=0.112613 s %t=0.225225 s %t=0.315315 s
    Scale bars 50 \textmu m.
    }
    \label{fig:fig1}
\end{figure}

%% Over one cardiac cycle, the CSF flow speed reaches up to
%% 28.0 \textmu m/s, dorsally in the posterior part of the middle
%% ventricle. This peak is attained at the time instant coinciding with
%% the time of peak caudal cardiac pulsatile flow,

@@

Interestingly, the distribution patterns differ between the
solutes. For the smallest diffusion coefficient, representing
extracellular vesicles, the advective transport by the CSF dominates diffusion: the
CSF flow structures carry over into the concentration field,
indicating that transport along the streamlines is more rapid than the
diffusion across the streamlines (\Crefpanels{fig:fig2}{b}). The
balance between diffusion and advection shifts with increasing
diffusion coefficient as the solute spreads more uniformly throughout
the ventricular geometry (\Crefpanels{fig:fig2}{b--d}). Notably, the
smaller solutes with higher diffusion coefficients spread more quickly
to distal regions. Computing Péclet numbers, we find that
$\mathrm{Pe}_1=664$, $\mathrm{Pe}_2=25$ and $\mathrm{Pe}_3=13$ for
$D_1$, $D_2$ and $D_3$, respectively. With all $\mathrm{Pe} > 1$, this
indicates that advection is the dominant transport mechanism on a
global scale.

@@

The CSF velocities are most strongly
affected when removing cilia in the dorsal diencephalic ventricle,
as the maximum velocity magnitude drops from \fixme{27.7} \textmu m/s to \fixme{8.0} \textmu m/s.
In vivo, these cilia generate the strongest flow at
this developmental stage of the fish~\cite{Olstad2019CiliaryDevelopment}.
In contrast, removing telencephalic and ventral diencephalic cilia has minimal
impact and results in maximum speeds of \fixme{27.7} \textmu m/s or \fixme{27.2} \textmu m/s, respectively.
Of the total tangential forces exerted by the cilia, ventral cilia in the middle ventricle
contribute with 34\%, while the dorsal cilia exert 52\% of the total tangential
force~(\Crefpanels{fig:fig5}{b}). The magnitude of the force contributions are thus
more similar than the difference in their impact on the maximum velocity magnitudes.

@@

Another study that used a force density to model ciliary motion is
Thouvenin~\emph{et al.}~\cite{Thouvenin2020OriginCanal}, where two-dimensional flow
in the central canal of zebrafish embryos was modeled using a cilia force density of
4000 N/$\mathrm{m^2}$. The resulting numerical velocity fields matched well with
in vivo measurements, where the maximum velocity magnitude was around 5 \textmu m/s
(recall that our model exhibits a maximum of 28.0 \textmu m/s).
The geometry they studied resembles a pipe, probably offering more flow resistance
than the open geometry considered in our work, which may be why they needed a larger
force density to accurately represent experimental results. In any case, it is
interesting to note that our and these two previous studies have estimated
cilia force densities that only differ by one order of magnitude.

@@

To evaluate the relative importance of advection and
diffusion as transport mechanisms, we consider the Péclet number
\begin{equation*}
    \mathrm{Pe} = \frac{\mathrm{diffusion \ timescale}}{\mathrm{advection \ timescale}}
    = \frac{t_d}{t_a} = \frac{L_p^2/D}{L_p/U_p} = \frac{L_p U_p}{D}.
\end{equation*}
Thus, if $\mathrm{Pe} > 1$, transport is dominated by advection,
and if $\mathrm{Pe} < 1$ transport is dominated by diffusion.
Here, $L_p$ and $U_p$ are characteristic length and velocity scales,
respectively, and $D$ is the solute diffusion coefficient.
In calculating a global Péclet number, we use $L_p=600$ \textmu m, the approximate length of the
ventricular geometry along the rostrocaudal ($x$-) axis.
We use a mean velocity $\overline{U}$ as the characteristic velocity $U_p$.
The mean velocity $\overline{U}$ is calculated by first averaging the velocity in space:
\begin{equation*}
    \overline{u}^2 = \frac{1}{|\Omega|}\int_{\Omega}\uu\cdot\uu\dx,
\end{equation*}
and then averaging $\overline{u}$ in time over one cardiac cycle to yield $\overline{U} = 2.4$ \textmu m/s.
To assess the assumption that inertial forces are negligible compared to
viscous forces, we also compute the Reynolds number
\begin{equation*}
    \mathrm{Re} = \frac{\mathrm{inertial \ forces}}{\mathrm{viscous \ forces}} = \frac{\rho U_r L_r}{\mu},
\end{equation*}
with $L_r = 110$ \textmu m (the height of the middle ventricle) and
$U_r = 28.0$ \textmu m/s (the maximum velocity magnitude)
as length and velocity scales characteristic of the flow.
